---
title: "Example: Cause-specific mortality rates"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
library(BayesGP)
library(TMB)
library(Matrix)
library(splines)
library(parallel)
library(ggplot2)
library(reshape2)
library(mixsqp)
library(tidyverse)
cpp_dir <- paste0(getwd(), "/code/cpp")
fig_dir <- paste0(getwd(), "/output/causes")
result_dir <- paste0(getwd(), "/output/causes")
function_dir <- paste0(getwd(), "/code/function")
data_dir <- paste0(getwd(), "/data")
source(paste0(function_dir, "/functions_fitting_Poisson_causes.R"))
compile(paste0(cpp_dir, "/Poisson_causes.cpp"))
compile(paste0(cpp_dir, "/Poisson_just_fixed_causes.cpp"))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Poisson_causes")))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Poisson_just_fixed_causes")))
num_cores <- 4
```


```{r}
#### Data:
deadFile = Pmisc::downloadIfOld("https://www150.statcan.gc.ca/n1/tbl/csv/13100810-eng.zip",
                                path = paste0(data_dir))
(deadFileCsv = deadFile[which.max(file.info(deadFile)$size)])
all_data = read.csv(deadFileCsv)

all_data$date = as.Date(as.character(all_data[[grep("DATE", names(all_data))]]))
all_data$province = gsub("[,].*", "", all_data$GEO)
all_data <- all_data %>% select(date, province, Cause.of.death..ICD.10.., VALUE)
```


Question of interest: For the five largest provinces in Canada, what are the mortality rates for the selected causes of death?
```{r}
selected_causes <- unique(all_data$Cause.of.death..ICD.10..)[c(2:11,16,18)]
provinces_selected <- c("Ontario", "Quebec", "British Columbia", "Alberta", "Manitoba")
provinces_data_selected <- all_data %>% filter(province %in% provinces_selected, date <= as.Date("2019-01-01"), `Cause.of.death..ICD.10..` %in% selected_causes)
colnames(provinces_data_selected)[4] <- "y"
```

To make the data less noisy, let's aggregate the each dataset to compute y being the average weekly death in each month:
```{r}
num_knots <- 30
region <- c(0,9)
provinces_data_selected$x <- (as.numeric(provinces_data_selected$date)/365) - (min(as.numeric(provinces_data_selected$date))/365)
datasets <- list()
# store all the datasets to study, at each pro and for each cause
for (i in 1:length(provinces_selected)){
  datasets[[i]] <- list()
  for (j in 1:length(selected_causes)){
    datasets[[i]][[j]] <- provinces_data_selected %>% 
      filter(province == provinces_selected[[i]], `Cause.of.death..ICD.10..` == selected_causes[[j]]) %>% 
      mutate(year = year(date), month = month(date)) %>% 
      group_by(year, month) %>% 
      reframe(y = mean(y), province = province, cause = selected_causes[[j]]) %>% 
      unique() %>%
      mutate(date = ceiling_date(as.Date(paste0(year, "-", month, "-01")), "month") - days(1)) %>%
      mutate(x = ((as.numeric(date)/365) - (min(as.numeric(provinces_data_selected$date))/365)))
  }
  names(datasets[[i]]) <- selected_causes
}
names(datasets) <- provinces_selected
```


```{r, eval=FALSE}
log_prec <- seq(-5,5, length.out = 10)
psd_vec <- sort(c(0,exp(-.5*log_prec)))
L_matrix_list <- c()
for (province in provinces_selected) {
  L_vecs <- mclapply(datasets[[province]], function(dataset) {
  compute_log_likelihood_sBspline_seq(
    dataset = dataset,
    num_knots = num_knots,
    psd_vector = psd_vec,
    pred_step = 5,
    betaprec = 0.0001,
    region = region
  )}, mc.cores = 6)
  L_matrix_list[[province]] <- do.call(rbind, L_vecs)
}
L_matrix <- do.call(rbind, L_matrix_list)
save(L_matrix_list, file = paste0(result_dir, "/L_matrix_list)
```

```{r, echo=FALSE}
log_prec <- seq(-5,5, length.out = 10)
psd_vec <- sort(c(0,exp(-.5*log_prec)))
load(paste0(result_dir, "/L_matrix_list.rda"))
L_matrix <- do.call(rbind, L_matrix_list)
```

```{r}
fit.sqp <- mixsqp(L = L_matrix, log = TRUE)
numiter <- nrow(fit.sqp$progress)
plot(1:numiter,fit.sqp$progress$objective,type = "b",
     pch = 20,lwd = 2,xlab = "SQP iteration",
     ylab = "objective",xaxp = c(1,numiter,numiter - 1))
prior_weight <- data.frame(psd = psd_vec, prior_weight = fit.sqp$x)
```

```{r, eval=FALSE}
num_weights <- sum(prior_weight$prior_weight != 0)
# Loop through each dataset and perform fitting
fitted_datasets <- list()
posterior_weights_matrix_list <- list()
for (province in provinces_selected) {
  fitted_datasets[[province]] <- list()
  posterior_weights_matrix_list[[province]] <- matrix(nrow = length(selected_causes), ncol = num_weights)
  for (j in 1:length(selected_causes)) {
    fit_result <- fit_sBspline_with_prior(
      y = datasets[[province]][[j]]$y,
      x = datasets[[province]][[j]]$x,
      num_knots = num_knots,
      pred_step = 5,
      betaprec = 0.0001,
      region = region,
      prior_weight = prior_weight
    )
    posterior_weights_matrix_list[[province]][j,] <- fit_result$posterior_weights[, "posterior_weight"]
    fitted_datasets[[province]][[j]] <- aggregate_fit_with_prior(x = datasets[[province]][[j]]$x, fit_results_with_prior = fit_result, original = TRUE)$summary_df
  }
}
save(fitted_datasets, posterior_weights_matrix_list, file = paste0(result_dir, "/fitted_datasets_and_pos_weights.rda"))
```

```{r, echo=FALSE}
num_weights <- sum(prior_weight$prior_weight != 0)
load(paste0(result_dir, "/fitted_datasets_and_pos_weights.rda"))
```


```{r}
## reduce posterior_weights_matrix_list to a single matrix
posterior_weights_matrix <- do.call(rbind, posterior_weights_matrix_list)
posterior_weights_df <- as.data.frame(posterior_weights_matrix)
posterior_weights_df$province <- rep(provinces_selected, each = length(selected_causes))
posterior_weights_df$cause <- rep(c(1:12), length(provinces_selected))
posterior_weights_df$id <- paste0(posterior_weights_df$province, "_", posterior_weights_df$cause)
colnames(posterior_weights_df)[1:4] <- prior_weight$psd[prior_weight$prior_weight != 0]
melted_data <- melt(posterior_weights_df[,c(1:4,7)], id.vars = "id")

ggplot(melted_data, aes(x = as.factor(id), y = value, fill = variable)) +
  geom_bar(stat = "identity") +
  labs(x = "Country", y = "Weight", fill = "PSD") +
  theme_minimal() +
  ggtitle("Structure Plot of Posterior Weights") +
  coord_flip()
```

```{r} 
# Normalize the data
wide_data <- scale(posterior_weights_matrix)
# Perform hierarchical clustering
d <- dist(wide_data, method = "euclidean")  # Distance matrix
fit <- hclust(d, method = "ward.D2")  # Clustering
clusters <- cutree(fit, k = 5)
posterior_weights_df$cluster <- clusters
melted_data <- melt(posterior_weights_df[,c(1:4,7,8)], id.vars = c("id", "cluster"))
melted_data <- melted_data %>% arrange(cluster)
ggplot(melted_data, aes(x = id, y = value, fill = variable)) +
  geom_bar(stat = "identity") +
  facet_wrap(~cluster, scales = "free_y") +  # Facet by cluster
  labs(x = "Country", y = "Weight", fill = "PSD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # Rotate x-axis labels for better readability
  ggtitle("Structure Plot of Posterior Weights by Cluster") +
  coord_flip()  
```


Let's visualize four example smoothing result for each cluster. 
For cluster 1:
```{r}
cluster1 <- posterior_weights_df %>% filter(cluster == 1)
cluster1$cause <- as.factor(cluster1$cause)
par(mfrow = c(3,3))
for (i in 1:9) {
  plot(fitted_datasets[[cluster1$province[i]]][[cluster1$cause[i]]]$x, datasets[[cluster1$province[i]]][[cluster1$cause[i]]]$y, type = "p", col = "black", xlab = "Time", ylab = "Mortality Rate", main = paste0("Province: ", cluster1$province[i], ", Cause: ", cluster1$cause[i]), cex = 0.5)
  lines(fitted_datasets[[cluster1$province[i]]][[cluster1$cause[i]]]$x, fitted_datasets[[cluster1$province[i]]][[cluster1$cause[i]]]$mean, col = "red")
}
```

For cluster 2:
```{r}
cluster2 <- posterior_weights_df %>% filter(cluster == 2)
cluster2$cause <- as.factor(cluster2$cause)
par(mfrow = c(2,2))
for (i in 1:4) {
  plot(fitted_datasets[[cluster2$province[i]]][[cluster2$cause[i]]]$x, datasets[[cluster2$province[i]]][[cluster2$cause[i]]]$y, type = "p", col = "black", xlab = "Time", ylab = "Mortality Rate", main = paste0("Province: ", cluster2$province[i], ", Cause: ", cluster2$cause[i]), cex = 0.5)
  lines(fitted_datasets[[cluster2$province[i]]][[cluster2$cause[i]]]$x, fitted_datasets[[cluster2$province[i]]][[cluster2$cause[i]]]$mean, col = "red")
}
```

For cluster 3:
```{r}
cluster3 <- posterior_weights_df %>% filter(cluster == 3)
cluster3$cause <- as.factor(cluster3$cause)
par(mfrow = c(2,2))
for (i in 1:4) {
  plot(fitted_datasets[[cluster3$province[i]]][[cluster3$cause[i]]]$x, datasets[[cluster3$province[i]]][[cluster3$cause[i]]]$y, type = "p", col = "black", xlab = "Time", ylab = "Mortality Rate", main = paste0("Province: ", cluster3$province[i], ", Cause: ", cluster3$cause[i]), cex = 0.5)
  lines(fitted_datasets[[cluster3$province[i]]][[cluster3$cause[i]]]$x, fitted_datasets[[cluster3$province[i]]][[cluster3$cause[i]]]$mean, col = "red")
}
```

For cluster 4:
```{r}
cluster4 <- posterior_weights_df %>% filter(cluster == 4)
cluster4$cause <- as.factor(cluster4$cause)
par(mfrow = c(2,2))
for (i in 1:4) {
  plot(fitted_datasets[[cluster4$province[i]]][[cluster4$cause[i]]]$x, datasets[[cluster4$province[i]]][[cluster4$cause[i]]]$y, type = "p", col = "black", xlab = "Time", ylab = "Mortality Rate", main = paste0("Province: ", cluster4$province[i], ", Cause: ", cluster4$cause[i]), cex = 0.5)
  lines(fitted_datasets[[cluster4$province[i]]][[cluster4$cause[i]]]$x, fitted_datasets[[cluster4$province[i]]][[cluster4$cause[i]]]$mean, col = "red")
}
```


For cluster 5:
```{r}
cluster5 <- posterior_weights_df %>% filter(cluster == 5)
cluster5$cause <- as.factor(cluster5$cause)
par(mfrow = c(2,1))
for (i in 1:2) {
  plot(fitted_datasets[[cluster5$province[i]]][[cluster5$cause[i]]]$x, datasets[[cluster5$province[i]]][[cluster5$cause[i]]]$y, type = "p", col = "black", xlab = "Time", ylab = "Mortality Rate", main = paste0("Province: ", cluster5$province[i], ", Cause: ", cluster5$cause[i]), cex = 0.5)
  lines(fitted_datasets[[cluster5$province[i]]][[cluster5$cause[i]]]$x, fitted_datasets[[cluster5$province[i]]][[cluster5$cause[i]]]$mean, col = "red")
}
```

