---
title: "Controlling based on BF"
author: "Ziang Zhang"
date: "2025-03-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Preprocessing $L$ matrix

Given a $L$ matrix of size $N \times (K + 1)$, we first collapse this matrix into a $N \times 2$ matrix $L^c$ through the following procedure:

- Compute the maximum likelihood estimate $\hat{\pi} = (\hat{\pi}_0, \hat{\pi}_1,..., \hat{\pi}_K)$.
- Reweight the alternative hypothesis weights $\hat{\pi}_1,..., \hat{\pi}_K$ into $\hat{\pi}_1^* = \hat{\pi}_1 / \sum_{k=1}^K \hat{\pi}_k$,..., $\hat{\pi}_K^* = \hat{\pi}_K / \sum_{k=1}^K \hat{\pi}_k$.
- Define $L^c_{i0} = L_{i0}$ and $L^c_{i1} = \sum_{k=1}^K L_{ik} \hat{\pi}_k^*$.

Then, we can compute the Bayes factor for each row of $L^c$ as:
$$\text{BF}_i = L^c_{i0}/L^c_{i1}.$$

Write a function for this:

```{r}
collapse_L <- function(L, log = F) {
  if (ncol(L) > 1) {
    pi_hat_star <- mixsqp::mixsqp(L = L[, -1, drop = FALSE],
                             log = log,
                             control = list(verbose = FALSE))$x
  } else{
    pi_hat_star <- rep(1, nrows(L))
  }
  L_c <- matrix(0, nrow = nrow(L), ncol = 2)
  L_c[, 1] <- L[, 1]
  L_c[, 2] <- (L[, -1, drop = FALSE] %*% pi_hat_star)
  return(L_c)
}
```

## Bayes factor based control

The BF based control procedure is as follows:

- For each cutoff $c$, compute $\mu(c) = \mathbb{E}(\text{BF} | \text{BF} \leq c) \approx \frac{\sum_{i=1}^N \mathbb{I}(\text{BF}_i \leq c)\text{BF}_i}{\sum_{i=1}^N \mathbb{I}(\text{BF}_i \leq c)}$, as well as $\hat{\pi}_0(c) = \frac{\sum_{i=1}^N \mathbb{I}(\text{BF}_i \leq c)}{N}$.
- Find $c^*$ such that $\mu(c^*) = 1$, return $\hat{\pi}_0(c^*)$ as the estimated for $\pi_0$.


Let's write a function for this:

```{r}
BF_control <- function(BF, plot = T) {
  BF_sorted <- sort(BF, decreasing = F)
  mu <- numeric(length(BF_sorted))
  pi0_hat <- numeric(length(BF_sorted))
  
  mu <- cumsum(BF_sorted) / seq_along(BF_sorted)
  pi0_hat <- seq_along(BF_sorted) / length(BF_sorted)
  
  if (max(mu) < 1) {
    pi0_hat_star <- 1
    return(list(mu = mu, pi0_hat = pi0_hat, pi0_hat_star = pi0_hat_star))
  }else{
    pi0_hat_star <- pi0_hat[mu >= 1][1]
  }
  
  if (plot) {
    
    par(mfrow = c(1, 2))
    
    hist(log(BF_sorted), breaks = 100, freq = T, xlab = "log-BF", main = "Histogram of log-BF")
    abline(v = log(BF_sorted)[mu >= 1][1], col = "red")
    
    plot(pi0_hat, mu, type = "l", xlab = "est pi0", ylab = "E(BF | BF <= c)", xlim=c(0,1), ylim = c(0,3))
    abline(h = 1, col = "red")

    par(mfrow = c(1, 1))
    
  }
  
  return(list(mu = mu, pi0_hat = pi0_hat, pi0_hat_star = pi0_hat_star))
  
}
```


## Example

```{r}
## simulate n observation from mixture of two normal distributions
n <- 30000
# Specify the true alternative hypothesis of N(0, true_sd)
true_sd <- 1
pi0_true <- 0.1
x <- c(rep(0, ceiling(n*pi0_true)), rnorm(n = ceiling(n*(1-pi0_true)), mean = 0, sd = true_sd))
## simulate y based on x
sebetahat <- 1
y <- x + rnorm(n, mean = 0, sd = sebetahat)
hist(y, breaks = 50)
## fit ash with a fitted alternative hypothesis of N(0, fit_sd)
fit_sd <- c(0.1,0.3,0.5)
ash_fit <- ashr::ash(betahat = y, sebetahat = sebetahat, mixcompdist = "normal", mixsd = fit_sd, nullweight = 1, outputlevel = 3)
ash_fit$fitted_g$pi[1]
L_mat <- ash_fit$fit_details$matrix_lik
## fit biased MoM
mean(apply(ash_fit$fit_details$matrix_lik, 1, which.max) == 1)
```


```{r}
L_c <- collapse_L(L_mat)
BF <- L_c[, 2] / L_c[, 1]
pi0_hat_star <- BF_control(BF)
# tail(pi0_hat_star$mu)
pi0_hat_star$pi0_hat_star
```

```{r}
BF1 <- BF[1:(pi0_true*n)]
BF2 <- BF[(pi0_true*n+1):n]

mean(BF1)
mean(BF2^(-1))

var(BF1) + (mean(BF1)^2)

mean(BF1^2)
mean(BF2)

mean(BF1^(-1))
mean(BF2^(-2))

par(mfrow = c(2, 2))
hist(log(BF1), breaks = 100, freq = F, xlab = "log-BF", main = "Histogram of log-BF: H0")
hist(log(BF2), breaks = 100, freq = F, xlab = "log-BF", main = "Histogram of log-BF: H1")
hist((BF1), breaks = 100, freq = F, xlab = "BF", main = "Histogram of BF: H0")
hist((BF2), breaks = 100, freq = F, xlab = "BF", main = "Histogram of BF: H1")
par(mfrow = c(1, 1))
```

```{r}
# ECDF of BF1
plot(ecdf(BF1), xlab = "BF", ylab = "F(BF)", main = "ECDF of BF: H0")
# Plot Markov Bound
markov_bound <- 1 - 1/seq(0, max(BF), by = 0.01)
lines(y = markov_bound, x = seq(0, max(BF), by = 0.01), col = "red")

# ECDF of BF2
plot(ecdf(BF2), xlab = "BF", ylab = "F(BF)", main = "ECDF of BF: H1")
```


## Why this procedure is conservative?

Let's see why the BF-based control gives conservative estimate of $\pi_0$.

Assume we have $N$ independent observations, with $N_0$ null hypotheses and $N_1$ alternative hypotheses. 

We say $i\in H_0$, if the $i$th observation is null, and $i\in H_1$, if the $i$th observation is alternative.

If $N$ is large, we know approximately that:
$$\pi_0 \approx N_0/N.$$
We also know that if $H_0$ is correctly specified,
$$\frac{\sum_{i \in H_0} \text{BF}_i}{N_0} \approx 1.$$

Suppose we define $\tilde{H}_0 = \{i\in[N]:\text{BF}_i\leq c^*\}$, where $c^*$ is the cutoff we found in the BF-based control procedure such that
$$\frac{\sum_{i \in \tilde{H}_0} \text{BF}_i}{\tilde{N}_0} \geq 1.$$

Then we claim:
$$\tilde{N}_0 \geq N_0.$$

To see this, assume that $|H_0 \cap \tilde{H}_0^C|= N_{0,1}$ and $|H_0 \cap \tilde{H}_0|= N_{0,0}$.
It is clear that $N_{0,1} + N_{0,0} = N_0$.

Further denote $|H_0^C \cap \tilde{H}_0|= N_{1,0}$, we know that $\hat{\pi}_0 = \frac{N_{0,0} + N_{1,0}}{N}$ and $\pi_0 \approx \frac{N_{0,0} + N_{0,1}}{N}$.

Therefore, we just need to show that $N_{1,0} \geq N_{0,1}$.

Using the fact that $\frac{\sum_{i \in H_0} \text{BF}_i}{N_0} = 1$, it is clear that:
\[
\frac{c^* N_{0,1} + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
\leq 
\frac{\sum_{i \in H_0 \cap \tilde{H}_0^C} \text{BF}_i + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
= 
\frac{\sum_{i \in H_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
= 1.
\]


Then using the fact that $\frac{\sum_{i \in \tilde{H}_0} \text{BF}_i}{\tilde{N}_0} \geq 1$, we can also show that:
\[
1 \leq \frac{\sum_{i\in H_0^C \cap \tilde{H}_0} \text{BF}_i + \sum_{i\in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{1,0} + N_{0,0}}
\leq \frac{c^* N_{1,0} + \sum_{i\in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{1,0} + N_{0,0}}
.
\]


Therefore, we know:
\[
\frac{c^* N_{0,1} + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
\leq 
\frac{c^* N_{1,0} + \sum_{i\in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{1,0} + N_{0,0}}.
\]


Define a function $$f(x) = \frac{c^*x + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{x + N_{0,0}}.$$

It is clear that $$f'(x) = \frac{c^* N_{0,0} - \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{(x + N_{0,0})^2}.$$

Since $\sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i \leq c^* N_{0,0}$, we know that $f'(x) \geq 0$ for all $x\geq 0$.

Therefore, we know that $f(x)$ is increasing in $x$. Hence $N_{1,0} \geq N_{0,1}$, and $\hat{\pi}_0 = (N_{1,0} + N_{0,0})/N  \geq (N_{0,1} + N_{0,0})/N =\pi_0$.


