---
title: "Controlling based on BF"
author: "Ziang Zhang"
date: "2025-03-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Preprocessing $L$ matrix

Given a $L$ matrix of size $N \times (K + 1)$, we first collapse this matrix into a $N \times 2$ matrix $L^c$ through the following procedure:

- Compute the maximum likelihood estimate $\hat{\pi} = (\hat{\pi}_0, \hat{\pi}_1,..., \hat{\pi}_K)$.
- Reweight the alternative hypothesis weights $\hat{\pi}_1,..., \hat{\pi}_K$ by $\hat{\pi}_j^* = \hat{\pi}_j / \sum_{k=1}^K \hat{\pi}_k$ for $j=1,...,K$.
- Define $L^c_{i0} = L_{i0}$ and $L^c_{i1} = \sum_{k=1}^K L_{ik} \hat{\pi}_k^*$.

Then, we can compute the Bayes factor for each row of $L^c$ as:
$$\text{BF}_i = L^c_{i0}/L^c_{i1}.$$

Write a function for this:

```{r}
collapse_L <- function(L, log = F) {
  if (ncol(L) > 1) {
    pi_hat_star <- mixsqp::mixsqp(L = L,
                             log = log,
                             control = list(verbose = FALSE))$x[-1]
    pi_hat_star <- pi_hat_star / sum(pi_hat_star)
  } else{
    pi_hat_star <- rep(1, nrows(L))
  }
  L_c <- matrix(0, nrow = nrow(L), ncol = 2)
  L_c[, 1] <- L[, 1]
  L_c[, 2] <- (L[, -1, drop = FALSE] %*% pi_hat_star)
  return(L_c)
}
```

## Bayes factor based control

The BF based control procedure is as follows:

- For each cutoff $c$, compute $\mu(c) = \mathbb{E}(\text{BF} | \text{BF} \leq c) \approx \frac{\sum_{i=1}^N \mathbb{I}(\text{BF}_i \leq c)\text{BF}_i}{\sum_{i=1}^N \mathbb{I}(\text{BF}_i \leq c)}$, as well as $\hat{\pi}_0(c) = \frac{\sum_{i=1}^N \mathbb{I}(\text{BF}_i \leq c)}{N}$.
- Find the smallest $c^*$ such that $\mu(c^*) \geq 1$, return $\hat{\pi}_0(c^*)$ as the estimate for $\pi_0$.
- Formally, $c^* = \inf\{c\in\mathbb{R}^+: \mu(c) \geq 1\}$.


Let's write a function for this:

```{r}
BF_control <- function(BF, plot = T) {
  BF_sorted <- sort(BF, decreasing = F)
  mu <- numeric(length(BF_sorted))
  pi0_hat <- numeric(length(BF_sorted))
  
  mu <- cumsum(BF_sorted) / seq_along(BF_sorted)
  pi0_hat <- seq_along(BF_sorted) / length(BF_sorted)
  
  if (max(mu) < 1) {
    pi0_hat_star <- 1
    return(list(mu = mu, pi0_hat = pi0_hat, pi0_hat_star = pi0_hat_star))
  }else{
    pi0_hat_star <- pi0_hat[mu >= 1][1]
  }
  
  if (plot) {
    
    par(mfrow = c(1, 2))
    
    hist(log(BF_sorted), breaks = 100, freq = T, xlab = "log-BF", main = "Histogram of log-BF")
    abline(v = log(BF_sorted)[mu >= 1][1], col = "red")
    
    plot(pi0_hat, mu, type = "l", xlab = "est pi0", ylab = "E(BF | BF <= c)", xlim=c(0,1), ylim = c(0,3))
    abline(h = 1, col = "red")

    par(mfrow = c(1, 1))
    
  }
  
  return(list(mu = mu, pi0_hat = pi0_hat, pi0_hat_star = pi0_hat_star))
  
}
```


## Example

```{r}
set.seed(1234)
## simulate n observation from mixture of two normal distributions
n <- 30000
# Specify the true alternative hypothesis of N(0, true_sd)
true_sd <- 1
pi0_true <- 0.3
x <- c(rep(0, ceiling(n*pi0_true)), rnorm(n = ceiling(n*(1-pi0_true)), mean = 0, sd = true_sd))
## simulate y based on x
sebetahat <- 0.5
y <- x + rnorm(n, mean = 0, sd = sebetahat)
hist(y, breaks = 50)
## fit ash with a fitted alternative hypothesis of N(0, fit_sd)
fit_sd <- c(0.1,0.3,0.5)
ash_fit <- ashr::ash(betahat = y, sebetahat = sebetahat, mixcompdist = "normal", mixsd = fit_sd, nullweight = 1, outputlevel = 3)
ash_fit$fitted_g$pi[1]
L_mat <- ash_fit$fit_details$matrix_lik
L_c <- collapse_L(L_mat)
BF <- L_c[, 2] / L_c[, 1]
pi0_hat_star <- BF_control(BF)
pi0_hat_star$pi0_hat_star
```

```{r}
BF1 <- BF[1:(pi0_true*n)]
# ECDF of BF1
plot(ecdf(BF1), xlab = "BF", ylab = "F(BF)", main = "ECDF of BF: Under null hypothesis")
# Plot Markov Bound
markov_bound <- 1 - 1/seq(0, 200, by = 0.01)
lines(y = markov_bound, x = seq(0, 200, by = 0.01), col = "red")
```


## Why this procedure is conservative?

Let's see why the BF-based control gives conservative estimate of $\pi_0$.

Assume we have $N$ independent observations, with $N_0$ from the null hypothesis and $N_1$ from the alternative hypothesis.

We say $i\in H_0$, if the $i$th observation is from the null, and $i\in H_1$, if the $i$th observation is from the alternative.

If $N$ is large, we know approximately (by LLN) that:
$$\pi_0 = N_0/N.$$
We also know by LLN that if $H_0$ is correctly specified, then (approximately)
$$\frac{\sum_{i \in H_0} \text{BF}_i}{N_0} = 1.$$

Let ${\tilde{N}_0}$ be the number of observations that are below the cutoff $c^*$, and $\tilde{H}_0$ be the set of indices such that $\text{BF}_i \leq c^*$.
Similarly, let $\tilde{N}_1$ be the number of observations that are above the cutoff $c^*$, and $\tilde{H}_1$ be the set of indices such that $\text{BF}_i > c^*$.

By construction we know
$$\frac{\sum_{i \in \tilde{H}_0} \text{BF}_i}{\tilde{N}_0} \geq 1.$$

The claim we try to prove is:
$$\tilde{N}_0 \geq N_0.$$

To see this, denote $N_{l,j} = |H_l \cap \tilde{H}_j^C|$ where $l,j \in \{0,1\}$.
It is clear that $N_{0,1} + N_{0,0} = N_0$, and $N_{1,0} + N_{0,0} = \tilde{N}_0$.

Hence $\hat{\pi}_0 = \frac{N_{0,0} + N_{1,0}}{N}$ and $\pi_0 = \frac{N_{0,0} + N_{0,1}}{N}$, and we just need to show that $N_{1,0} \geq N_{0,1}$.

Using the fact that $\frac{\sum_{i \in H_0} \text{BF}_i}{N_0} = 1$, it is clear that:
\[
\frac{c^* N_{0,1} + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
\leq 
\frac{\sum_{i \in H_0 \cap \tilde{H}_0^C} \text{BF}_i + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
= 
\frac{\sum_{i \in H_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
= 1.
\]


Then using the fact that $\frac{\sum_{i \in \tilde{H}_0} \text{BF}_i}{\tilde{N}_0} \geq 1$, we can also show that:
\[
1 \leq \frac{\sum_{i\in H_0^C \cap \tilde{H}_0} \text{BF}_i + \sum_{i\in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{1,0} + N_{0,0}}
\leq \frac{c^* N_{1,0} + \sum_{i\in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{1,0} + N_{0,0}}
.
\]


Therefore, we know:
\[
\frac{c^* N_{0,1} + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{0,1} + N_{0,0}} 
\leq 1 \leq
\frac{c^* N_{1,0} + \sum_{i\in H_0 \cap \tilde{H}_0} \text{BF}_i}{N_{1,0} + N_{0,0}}.
\]


Define a function $$f(x) = \frac{c^*x + \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{x + N_{0,0}}.$$

It is clear that $$f'(x) = \frac{c^* N_{0,0} - \sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i}{(x + N_{0,0})^2}.$$

Since $\sum_{i \in H_0 \cap \tilde{H}_0} \text{BF}_i \leq c^* N_{0,0}$, we know that $f'(x) \geq 0$ for all $x\geq 0$.

Therefore, we know that $f(x)$ is increasing in $x$. Hence $N_{1,0} \geq N_{0,1}$, and $\hat{\pi}_0 = (N_{1,0} + N_{0,0})/N  \geq (N_{0,1} + N_{0,0})/N =\pi_0$.


