---
title: "Iyer"
author: "Ziang Zhang"
date: "2024-12-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r message=FALSE, warning=FALSE}
library(BayesGP)
library(TMB)
library(Matrix)
library(splines)
library(parallel)
library(ggplot2)
library(reshape2)
library(mixsqp)
library(tidyverse)
library(mashr)
cpp_dir <- paste0(getwd(), "/code/cpp")
function_dir <- paste0(getwd(), "/code/function")
result_dir <- paste0(getwd(), "/output/Iyer")
data_dir <- paste0(getwd(), "/data/Iyer")
source(paste0(function_dir, "/functions_fitting_Gaussian.R"))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Gaussian_theta_known")))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Gaussian_just_fixed")))
load(paste0(data_dir, "/Iyer_expression.rda"))
load(paste0(data_dir, "/sigmaE_Iyer.rda"))
```

Create a list of 517 datasets, one for each gene:

```{r}
all_gene <- Iyer_expression$`Cluster order`
datasets <- lapply(1:517, function(i) {
  data.frame(
    x = c(15, 30, 60, 120, 240, 360, 480, 720, 960, 1200, 1440)/60,
    y = as.numeric(Iyer_expression[i,6:16]),
    sigmaE = sigmaE[i]
  )
})
```

Doing EB to learn the mixture weights:

```{r eval=FALSE}
set.seed(123)
p_vec <- 2
log_prec <- seq(5,15,length=30)
psd_vec <- c(1/exp(.5*log_prec), 0)

# Using the FEM:
L_vecs <- mclapply(datasets, function(dataset) {
  compute_log_likelihood_ospline_seq2(
    x = dataset$x,
    y = dataset$y,
    p = p_vec,
    num_knots = 10,
    psd_iwp_vector = psd_vec,
    pred_step = 15/60,
    betaprec = 0.001,
    sd_gaussian = 0.2207696
  )
}, mc.cores = 1)
L_matrix <- do.call(rbind, L_vecs)

save(L_matrix, file = paste0(result_dir, "/L_matrix.rda"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
p_vec <- 2
log_prec <- seq(5,15,length=30)
psd_vec <- c(1/exp(.5*log_prec), 0)
load(paste0(result_dir, "/L_matrix.rda"))
```


Compute the Posterior:

```{r}
fit.sqp <- mixsqp(L = L_matrix, log = TRUE)
numiter <- nrow(fit.sqp$progress)
plot(1:numiter,fit.sqp$progress$objective,type = "b",
     pch = 20,lwd = 2,xlab = "SQP iteration",
     ylab = "objective",xaxp = c(1,numiter,numiter - 1))
prior_weight <- data.frame(p = rep(p_vec, each = length(psd_vec)), psd_iwp = psd_vec, prior_weight = fit.sqp$x)

posterior_matrix <- matrix(0, nrow = nrow(L_matrix), ncol = ncol(L_matrix))
for(i in 1:nrow(L_matrix)){
  posterior_matrix[i,] <- exp(L_matrix[i,] - max(L_matrix[i,]) + log(fit.sqp$x))
  posterior_matrix[i,] <- posterior_matrix[i,]/sum(posterior_matrix[i,])
}
colnames(posterior_matrix) <- paste0(p_vec,"_",psd_vec)
posterior_weights_matrix <- posterior_matrix
```

Visualize the posterior weights:

```{r}
mean_psd <- apply(posterior_weights_matrix, 1, function(x) {
  weighted_psd <- as.numeric(sub("*._", "", colnames(posterior_weights_matrix)))
  sum(x * weighted_psd)
})
sorted_index <- order(mean_psd)
sorted_posterior_weights_matrix <- posterior_weights_matrix[sorted_index, ]
sorted_posterior_weights_df <- as.data.frame(sorted_posterior_weights_matrix)
sorted_posterior_weights_df$id <- all_gene[sorted_index]
melted_data_sorted <- melt(sorted_posterior_weights_df, id.vars = "id")
melted_data_sorted$variable2 <- sub("_.*", "", melted_data_sorted$variable)
melted_data_sorted$variable3 <- (round(as.numeric(sub("*._", "", melted_data_sorted$variable)), 3))


ggplot(melted_data_sorted, aes(x = factor(id, levels = sorted_posterior_weights_df$id), y = value, fill = variable3)) +
  geom_bar(stat = "identity") +
  labs(x = "Gene", y = "Weight", fill = "PSD") +
  ggtitle("Structure Plot of Posterior Weights") +
  coord_flip() +
  scale_fill_gradient(low = "white", high = "blue") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),  
    panel.background = element_rect(fill = "white", colour = "grey"),
    plot.background = element_rect(fill = "white", colour = NA) 
  )
```

Let's group them into `num_clusters` groups, based on the quantiles of the mean PSD:
```{r message=FALSE, warning=FALSE}
mean_psd_ordered <- mean_psd[sorted_index]
quantiles <- c(0, 0.004, 0.01, 0.025, 0.05)
quantile_groups <- cut(mean_psd_ordered, breaks = quantiles, 
                       include.lowest = TRUE, labels = FALSE)
table(quantile_groups)
num_clusters <- length(unique(quantile_groups))

sorted_posterior_weights_df$cluster <- quantile_groups
melted_data_quantiles <- melt(sorted_posterior_weights_df, id.vars = c("id", "cluster"))
melted_data_quantiles$variable2 <- sub("_.*", "", melted_data_quantiles$variable)
melted_data_quantiles$variable3 <- round(as.numeric(sub(".*_", "", melted_data_quantiles$variable)), 3)
melted_data_quantiles <- melted_data_quantiles %>% arrange(cluster)

# Ensure the 'id' factor is ordered according to the sorted mean_psd
melted_data_quantiles$id <- factor(melted_data_quantiles$id, levels = sorted_posterior_weights_df$id)

# Plotting with facets by cluster and proper sorting of 'id'
ggplot(melted_data_quantiles, aes(x = factor(id, levels = sorted_posterior_weights_df$id), y = value, fill = variable3)) +
  geom_bar(stat = "identity") +
  labs(x = "Gene", y = "Weight", fill = "PSD Level") +
  ggtitle("Structure Plot of Posterior Weights by Quantile Group") +
  coord_flip() +
  scale_fill_gradient(low = "white", high = "blue") +
  facet_wrap(~cluster, scales = "free_y") +  # Faceting by cluster while keeping order
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.background = element_rect(fill = "white", colour = "grey"),
    plot.background = element_rect(fill = "white", colour = NA),
    strip.background = element_rect(fill = "lightblue", colour = "deepskyblue", size = 1),
    strip.text = element_text(face = "bold")
  )
```

```{r}
# Automatically extract gene IDs for each cluster
num_genes_per_cluster <- 6
genes_per_cluster <- list()
for (i in 1:num_clusters) {
  genes_per_cluster[[i]] <- as.numeric(sorted_posterior_weights_df$id[sorted_posterior_weights_df$cluster == i][1:num_genes_per_cluster])
}
```


Take a look at the inferred trajectories:


```{r, eval=FALSE}
num_datasets <- length(datasets)
num_weights <- sum(prior_weight$prior_weight != 0)
posterior_weights_matrix <- matrix(nrow = num_datasets, ncol = num_weights)

# Loop through each dataset and perform fitting
fitted_datasets <- list()

# Using FEM:
for (i in seq_along(datasets)) {
  dataset <- datasets[[i]]
  fit_result_final <- fit_ospline_with_prior2(
    num_cores = 1,
    x = dataset$x,
    y = dataset$y,
    num_knots = 10,
    prior_weight = prior_weight,
    betaprec = 0.001,
    sd_gaussian = 0.2207696,
    pred_step = 15/60
  )
  fitted_datasets[[i]] <- aggregate_fit_with_prior(x = dataset$x, fit_results_with_prior = fit_result_final)$summary_df
}
save(fitted_datasets, file = paste0(result_dir, "/fitted_datasets.rda"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
num_datasets <- length(datasets)
num_weights <- sum(prior_weight$prior_weight != 0)
posterior_weights_matrix <- matrix(nrow = num_datasets, ncol = num_weights)
load(paste0(result_dir, "/fitted_datasets.rda"))
```

Show some selected results from each of the four clusters:

```{r}
par(mfrow = c(3, 2), mar = c(2, 2, 1, 1))  
for (i in genes_per_cluster[[1]]) {
  plot(
    datasets[[i]]$x,
    datasets[[i]]$y,
    type = 'p',
    main = paste("Dataset", i),
    xlab = "x",
    ylab = "y",
    ylim = c(-2,5),
    cex = 0.2,
    cex.main = 0.8,
    cex.lab = 0.7,
    cex.axis = 0.7
  )
  lines(fitted_datasets[[i]]$x,
        fitted_datasets[[i]]$mean,
        col = "blue",
        lwd = 1.2)
  polygon(
    c(fitted_datasets[[i]]$x, rev(fitted_datasets[[i]]$x)),
    c(fitted_datasets[[i]]$lower, rev(fitted_datasets[[i]]$upper)),
    col = rgb(0, 0, 1, 0.2),
    border = NA
  )
}
par(mfrow = c(1, 1))
```

```{r}
par(mfrow = c(3, 2), mar = c(2, 2, 1, 1))  
for (i in genes_per_cluster[[2]]) {
  plot(
    datasets[[i]]$x,
    datasets[[i]]$y,
    type = 'p',
    main = paste("Dataset", i),
    xlab = "x",
    ylab = "y",
    ylim = c(-2,5),
    cex = 0.2,
    cex.main = 0.8,
    cex.lab = 0.7,
    cex.axis = 0.7
  )
  lines(fitted_datasets[[i]]$x,
        fitted_datasets[[i]]$mean,
        col = "blue",
        lwd = 1.2)
  polygon(
    c(fitted_datasets[[i]]$x, rev(fitted_datasets[[i]]$x)),
    c(fitted_datasets[[i]]$lower, rev(fitted_datasets[[i]]$upper)),
    col = rgb(0, 0, 1, 0.2),
    border = NA
  )
}
par(mfrow = c(1, 1))
```

```{r}
par(mfrow = c(3, 2), mar = c(2, 2, 1, 1))
for (i in genes_per_cluster[[3]]) {
  plot(
    datasets[[i]]$x,
    datasets[[i]]$y,
    type = 'p',
    main = paste("Dataset", i),
    xlab = "x",
    ylab = "y",
    ylim = c(-2,5),
    cex = 0.2,
    cex.main = 0.8,
    cex.lab = 0.7,
    cex.axis = 0.7
  )
  lines(fitted_datasets[[i]]$x,
        fitted_datasets[[i]]$mean,
        col = "blue",
        lwd = 1.2)
  polygon(
    c(fitted_datasets[[i]]$x, rev(fitted_datasets[[i]]$x)),
    c(fitted_datasets[[i]]$lower, rev(fitted_datasets[[i]]$upper)),
    col = rgb(0, 0, 1, 0.2),
    border = NA
  )
}
par(mfrow = c(1, 1))
```

```{r}
par(mfrow = c(3, 2), mar = c(2, 2, 1, 1))
for (i in genes_per_cluster[[4]]) {
  plot(
    datasets[[i]]$x,
    datasets[[i]]$y,
    type = 'p',
    main = paste("Dataset", i),
    xlab = "x",
    ylab = "y",
    ylim = c(-2,5),
    cex = 0.2,
    cex.main = 0.8,
    cex.lab = 0.7,
    cex.axis = 0.7
  )
  lines(fitted_datasets[[i]]$x,
        fitted_datasets[[i]]$mean,
        col = "blue",
        lwd = 1.2)
  polygon(
    c(fitted_datasets[[i]]$x, rev(fitted_datasets[[i]]$x)),
    c(fitted_datasets[[i]]$lower, rev(fitted_datasets[[i]]$upper)),
    col = rgb(0, 0, 1, 0.2),
    border = NA
  )
}
par(mfrow = c(1, 1))
```


Produce a plot where the posterior mean of each gene from the given cluster is plotted:

```{r}
all_genes_per_cluster <- list()
for (i in 1:num_clusters) {
  all_genes_per_cluster[[i]] <- as.numeric(sorted_posterior_weights_df$id[sorted_posterior_weights_df$cluster == i])
}
```


```{r}
# for cluster 1
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
# initialize the plot
plot(
  fitted_datasets[[all_genes_per_cluster[[1]][1]]]$x,
  fitted_datasets[[all_genes_per_cluster[[1]][1]]]$mean,
  type = 'l',
  # choose a very transparent color
  col = rgb(0, 0, 1, 0.1),
  xlab = "x",
  ylab = "y",
  ylim = c(-2,5),
  cex = 0.2,
  cex.main = 0.8,
  cex.lab = 0.7,
  cex.axis = 0.7,
)
for (i in all_genes_per_cluster[[1]]) {
  lines(
    fitted_datasets[[i]]$x,
    fitted_datasets[[i]]$mean,
    col = rgb(0, 0, 1, 0.1),
    lwd = 1.5
  )
}
```

```{r}
# for cluster 2
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
# initialize the plot
plot(
  fitted_datasets[[all_genes_per_cluster[[2]][1]]]$x,
  fitted_datasets[[all_genes_per_cluster[[2]][1]]]$mean,
  type = 'l',
  # choose a very transparent color
  col = rgb(0, 0, 1, 0.1),
  xlab = "x",
  ylab = "y",
  ylim = c(-2,5),
  cex = 0.2,
  cex.main = 0.8,
  cex.lab = 0.7,
  cex.axis = 0.7,
)
for (i in all_genes_per_cluster[[2]]) {
  lines(
    fitted_datasets[[i]]$x,
    fitted_datasets[[i]]$mean,
    col = rgb(0, 0, 1, 0.2),
    lwd = 1.5
  )
}
```

```{r}
# for cluster 3
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
# initialize the plot
plot(
  fitted_datasets[[all_genes_per_cluster[[3]][1]]]$x,
  fitted_datasets[[all_genes_per_cluster[[3]][1]]]$mean,
  type = 'l',
  # choose a very transparent color
  col = rgb(0, 0, 1, 0.1),
  xlab = "x",
  ylab = "y",
  ylim = c(-2,5),
  cex = 0.2,
  cex.main = 0.8,
  cex.lab = 0.7,
  cex.axis = 0.7,
)
for (i in all_genes_per_cluster[[3]]) {
  lines(
    fitted_datasets[[i]]$x,
    fitted_datasets[[i]]$mean,
    col = rgb(0, 0, 1, 0.2),
    lwd = 1.5
  )
}
```

```{r}
# for cluster 4
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
# initialize the plot
plot(
  fitted_datasets[[all_genes_per_cluster[[4]][1]]]$x,
  fitted_datasets[[all_genes_per_cluster[[4]][1]]]$mean,
  type = 'l',
  # choose a very transparent color
  col = rgb(0, 0, 1, 0.1),
  xlab = "x",
  ylab = "y",
  ylim = c(-2,5),
  cex = 0.2,
  cex.main = 0.8,
  cex.lab = 0.7,
  cex.axis = 0.7,
)
for (i in all_genes_per_cluster[[4]]) {
  lines(
    fitted_datasets[[i]]$x,
    fitted_datasets[[i]]$mean,
    col = rgb(0, 0, 1, 0.2),
    lwd = 1.5
  )
}
```

