---
title: "Circadian Expression Detection"
author: "Ziang Zhang"
date: "2024-06-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(BayesGP)
library(TMB)
library(Matrix)
library(splines)
library(parallel)
library(ggplot2)
library(reshape2)
library(mixsqp)
library(tidyverse)
cpp_dir <- paste0(getwd(), "/code/cpp")
fig_dir <- paste0(getwd(), "/output/circadian_expression")
result_dir <- paste0(getwd(), "/output/circadian_expression")
function_dir <- paste0(getwd(), "/code/function")
data_dir <- paste0(getwd(), "/data/circadian_expression")
source(paste0(function_dir, "/functions_fitting_Poisson_circadian_expression.R"))
compile(paste0(cpp_dir, "/Poisson_expression.cpp"))
compile(paste0(cpp_dir, "/Poisson_just_fixed_expression.cpp"))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Poisson_expression")))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Poisson_just_fixed_expression")))
num_cores <- 1
```



```{r}
# Load data
load(file = paste0(data_dir, "/circ_data.rda"))
beta_prec <- 1e-10
k <- 30
p <- 4
psd_vec <- seq(0, 3, by = 0.1)
```

## Focus on Group C first:

```{r eval = FALSE}
datasets <- list()
all_gene <- unique(gene_exp_data_c$gene_id)
for (gene in all_gene) {
  selected_row <- gene_exp_data_c %>% filter(gene_id == gene)
  datasets[[gene]] <- data.frame(x = seq(0, 44, by = 4))/24
  datasets[[gene]]$y <- as.numeric(selected_row[-c(1,2)])
}
save(datasets, file = paste0(result_dir, "/datasets.rda"))
```

```{r}
datasets <- list()
all_gene <- unique(gene_exp_data_c$gene_id)
load(paste0(result_dir, "/datasets.rda"))
```

To ensure the accuracy of the marginal likelihood approximation, we should filter out the genes with extremely small number of counts

```{r}
datasets_original <- datasets
datasets <- list()
for (gene in all_gene) {
  if(min(datasets_original[[gene]]$y) > 100) {
    datasets[[gene]] <- datasets_original[[gene]]
    datasets[[gene]]$y <- round(datasets[[gene]]$y)
  }
}
# update all_gene
all_gene <- datasets %>% names()
```

Compute the library size:

```{r}
size_vec <- numeric(length = nrow(datasets[[1]]))
for (i in 1:length(size_vec)) {
  all_counts <- unlist(lapply(datasets, function(x) {x$y[i]}))
  size_vec[i] <- sum(all_counts)
}
log_size_vec <- log(size_vec)
```

Visualize 16 of the datasets:

```{r}
par(mfrow = c(4, 4))
for (i in 1:16) {
  gene <- all_gene[i]
  plot(datasets[[gene]]$x, datasets[[gene]]$y, type = "o", 
       col = "blue", lwd = 2, xlab = "Time", ylab = "Expression")
  abline(v = 1, lty = 2)
}
par(mfrow = c(1, 1))
```

Compute the L matrix

```{r eval = FALSE}
## set up a progress bar
L_vecs <- vector("list", length(datasets))
pb <- txtProgressBar(min = 0, max = length(datasets), style = 3)
for (i in 1:length(datasets)) {
  setTxtProgressBar(pb, i)
  data <- datasets[[i]]

  # time_series <- ts(data$y, frequency = 6)
  # 
  # # do not show the plot
  # dd <- spectrum(time_series, method = c("pgram"), plot = F, pad = 4)
  # possible_period <- round((1/dd$freq)[(1/dd$freq) >= 0.8 & (1/dd$freq) <= 1.2], 3)
  # 
  # period_df <- data.frame(period = round(1/dd$freq, 3), spec = dd$spec)
  # period_df <- period_df %>% filter(period %in% possible_period)
  # period <- period_df$period[which.max(period_df$spec)]
  
  # due to the lack of data, fixed at 24 hrs!
  period <- 1
  
  like_vec1 <- compute_log_likelihood_sBspline_seq(dataset = data, 
                                    period = period, 
                                    num_knots = k, 
                                    psd_sgp_vector = psd_vec,
                                    pred_step = period,
                                    betaprec = beta_prec, 
                                    log_lib_size = log_size_vec, 
                                    m = 1)
  like_vec2 <- compute_log_likelihood_ospline_seq(dataset = data, 
                                    p = p, 
                                    num_knots = k, 
                                    psd_iwp_vector = psd_vec,
                                    pred_step = period,
                                    betaprec = beta_prec, 
                                    log_lib_size = log_size_vec)
  
  L_vecs[[i]] <- c(like_vec1, like_vec2)
}
L_matrix <- do.call(rbind, L_vecs)
save(L_matrix, file = paste0(result_dir, "/L_matrix.rda"))
```

Store the periodicity vector:
```{r}
load(paste0(result_dir, "/L_matrix.rda"))
period_vec <- numeric(length = length(datasets))
pb <- txtProgressBar(min = 0, max = length(datasets), style = 3)
for (i in 1:length(datasets)) {
  setTxtProgressBar(pb, i)
  data <- datasets[[i]]
  time_series <- ts(data$y, frequency = 6)
  
  # do not show the plot
  dd <- spectrum(time_series, method = c("pgram"), plot = F, pad = 4)
  possible_period <- round((1/dd$freq)[(1/dd$freq) >= 0.8 & (1/dd$freq) <= 1.2], 3)

  period_df <- data.frame(period = round(1/dd$freq, 3), spec = dd$spec)
  period_df <- period_df %>% filter(period %in% possible_period)
  the_period <- period_df$period[which.max(period_df$spec)]
  # period_vec[i] <- the_period
  # due to the lack of data, fixed at 24 hrs!
  period_vec[i] <- 1
}
```


```{r}
# remove the rows with NaN
removed_index <- which(apply(L_matrix, 1, function(x) {any(is.nan(x))}))
L_matrix <- L_matrix[-removed_index, ]

fit.sqp <- mixsqp(L = L_matrix, log = TRUE)
numiter <- nrow(fit.sqp$progress)
plot(1:numiter,fit.sqp$progress$objective,type = "b",
     pch = 20,lwd = 2,xlab = "SQP iteration",
     ylab = "objective",xaxp = c(1,numiter,numiter - 1))

# create prior weight 
prior_weight <- data.frame(name = paste0(rep(c("sgp_", "iwp_"), each = length(psd_vec)), psd_vec), prior_weight = fit.sqp$x)
prior_weight

# sum the prior weight corresponding to sgp
sum(fit.sqp$x[grepl("sgp_", prior_weight$name)])
```


Compute the posterior:

```{r}
# Compute the posterior using the likelihood and prior
posterior_matrix <- matrix(0, nrow = nrow(L_matrix), ncol = ncol(L_matrix))
for(i in 1:nrow(L_matrix)){
  posterior_matrix[i,] <- exp(L_matrix[i,] - max(L_matrix[i,]) + log(fit.sqp$x))
  posterior_matrix[i,] <- posterior_matrix[i,]/sum(posterior_matrix[i,])
}
```

```{r}
colnames(posterior_matrix) <- paste0(rep(c("sgp_", "iwp_"), each = length(psd_vec)), psd_vec)
```

The posterior probability of being in class sgp (summing all sgp columns):

```{r}
posterior_prob <- rowSums(posterior_matrix[, grepl("sgp_", colnames(posterior_matrix))])
head(posterior_prob)
```


The local FDR for testing periodicity:

```{r}
lfdr <- 1 - posterior_prob
lfdr <- ifelse(lfdr < 0, 0, lfdr) # set to 0 if it is negative due to numerical error
lfdr <- ifelse(lfdr > 1, 1, lfdr) # set to 1 if it is greater than 1 due to numerical error

fdr_df <- data.frame(ID = 1:length(lfdr), fdr = lfdr, NAMES = all_gene[-removed_index], period = period_vec[-removed_index])
fdr_df <- fdr_df[order(fdr_df$fdr), ] # ordering it
fdr_df$cumulative_fdr <- cumsum(fdr_df$fdr)/seq_along(fdr_df$fdr)
fdr_df$rank <- 1:length(lfdr)
```


Plot the FDR curve:

```{r}
ggplot(fdr_df, aes(x = 1:length(lfdr), y = cumulative_fdr)) +
  geom_point() +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "purple") +
  labs(x = "Ordered Units", y = "Cumulative FDR", col = "Type") +
  theme_minimal() +
  ggtitle("FASH:Cumulative FDR Plot")
```


Visualize the leading genes in the non-periodic and the periodic groups:

```{r}
non_periodic_genes <- fdr_df %>% filter(cumulative_fdr > 0.05) %>% pull(NAMES) %>% tail(16)

periodic_genes <- fdr_df %>% filter(cumulative_fdr <= 0.05) %>% pull(NAMES) %>% head(16)

# visualize 4 of the non-periodic genes
par(mfrow = c(2, 2))
for (i in 1:4) {
  gene <- non_periodic_genes[i]
  plot(datasets[[gene]]$x, datasets[[gene]]$y, type = "o", 
       col = "blue", lwd = 2, xlab = "Time", ylab = "Expression")
  the_period <- fdr_df %>% filter(NAMES == gene) %>% pull(period)
  abline(v = seq(0, 2, by = the_period), lty = 2)
}
par(mfrow = c(1, 1))

# visualize 4 of the periodic genes
par(mfrow = c(2, 2))
for (i in 1:4) {
  gene <- periodic_genes[i]
  plot(datasets[[gene]]$x, datasets[[gene]]$y, type = "o", 
       col = "blue", lwd = 2, xlab = "Time", ylab = "Expression")
  the_period <- fdr_df %>% filter(NAMES == gene) %>% pull(period)
  abline(v = seq(0, 2, by = the_period), lty = 2)
}
par(mfrow = c(1, 1))
```


The periodic genes seem to have similar pattern of going up and down:
```{r}
par(mfrow = c(2, 2))
for (i in 1:4) {
  gene <- periodic_genes[i]
  data <- datasets[[gene]]
  data$change <- sign(c(diff(data$y), NA))
  plot(data$x, data$change, type = "o", 
       col = "blue", lwd = 2, xlab = "Time", ylab = "Change")
  the_period <- fdr_df %>% filter(NAMES == gene) %>% pull(period)
  abline(v = seq(0, 2, by = the_period), lty = 2)
}
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))  # Set up a 2x2 plot layout
for (i in 1:4) {
  gene <- periodic_genes[i]
  data <- datasets[[gene]]
  data$change <- sign(c(diff(data$y), NA))
  data <- data[!is.na(data$change), ]  # Remove rows with NA
  period_length <- fdr_df %>% filter(NAMES == gene) %>% pull(period)
  period1 <- data[data$x < period_length, ]
  period1$folded_x <- round(period1$x, 3) %% period_length
  period2 <- data[data$x >= period_length & data$x <= 2*period_length, ]
  period2$folded_x <- round(period2$x, 3) %% period_length
  
  plot(period1$folded_x, period1$change, type = "o", col = "blue", lwd = 2, 
       xlab = "Folded Time", ylab = "Change", main = paste("Gene", gene), ylim = range(c(period1$change, period2$change)), lty = "dashed")
  
  points(period2$folded_x, period2$change, type = "o", col = "red", lwd = 2, lty = "dotted")  # Plot the second period in red
  
}
par(mfrow = c(1, 1))  # Reset the plotting layout

```

Let's look at the non-periodic genes for comparison:

```{r}
par(mfrow = c(2, 2))
for (i in 1:4) {
  gene <- non_periodic_genes[i]
  data <- datasets[[gene]]
  data$change <- sign(c(diff(data$y), NA))
  plot(data$x, data$change, type = "o", 
       col = "blue", lwd = 2, xlab = "Time", ylab = "Change")
  the_period <- fdr_df %>% filter(NAMES == gene) %>% pull(period)
  abline(v = seq(0, 2, by = the_period), lty = 2)
}
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))  # Set up a 2x2 plot layout
for (i in 1:4) {
  gene <- non_periodic_genes[i]
  data <- datasets[[gene]]
  data$change <- sign(c(diff(data$y), NA))
  data <- data[!is.na(data$change), ]  # Remove rows with NA
  period_length <- fdr_df %>% filter(NAMES == gene) %>% pull(period)
  period1 <- data[data$x < period_length, ]
  period1$folded_x <- round(period1$x, 3) %% period_length
  period2 <- data[data$x >= period_length & data$x <= 2*period_length, ]
  period2$folded_x <- round(period2$x, 3) %% period_length
  
  plot(period1$folded_x, period1$change, type = "o", col = "blue", lwd = 2, 
       xlab = "Folded Time", ylab = "Change", main = paste("Gene", gene), ylim = range(c(period1$change, period2$change)), lty = "dashed")
  
  points(period2$folded_x, period2$change, type = "o", col = "red", lwd = 2, lty = "dotted")  # Plot the second period in red
  
  abline(v = 1, lty = 2)  # Add a vertical line at the period boundary
}
par(mfrow = c(1, 1))  # Reset the plotting layout

```


<!-- Take a look at the fitted functions: -->

<!-- ```{r} -->
<!-- # visualize 4 of the non-periodic genes -->
<!-- ggplot_list <- list() -->
<!-- for (i in 1:4) { -->
<!--   gene <- non_periodic_genes[i] -->
<!--   data <- datasets[[gene]] -->
<!--   # compute the prior weight -->
<!--   prior_weight_iwp <- prior_weight %>% filter(grepl("iwp_", name)) -->
<!--   # renormalize the prior weight -->
<!--   prior_weight_iwp$prior_weight <- prior_weight_iwp$prior_weight/sum(prior_weight_iwp$prior_weight) -->
<!--   prior_weight_iwp$name <- psd_vec -->
<!--   prior_weight_iwp <- rename(prior_weight_iwp, psd_iwp = name) -->
<!--   # fit the model -->
<!--   fit <- fit_ospline_with_prior(data = data, p = p, num_knots = k, betaprec = beta_prec, log_lib_size = log_size_vec, prior_weight = prior_weight_iwp, pred_step = 1) -->
<!--   ggplot_list[[i]] <- visualize_fit_with_prior(x = data$x, fit_result = fit) -->
<!-- } -->

<!-- # show the ggplot in a 2 x 2 grid -->
<!-- gridExtra::grid.arrange(grobs = ggplot_list, ncol = 2) -->

<!-- # visualize 4 of the periodic genes -->
<!-- ggplot_list <- list() -->
<!-- for (i in 1:4) { -->
<!--   gene <- periodic_genes[i] -->
<!--   data <- datasets[[gene]] -->
<!--   # compute the prior weight -->
<!--   prior_weight_sgp <- prior_weight %>% filter(grepl("sgp_", name)) -->
<!--   # renormalize the prior weight -->
<!--   prior_weight_sgp$prior_weight <- prior_weight_sgp$prior_weight/sum(prior_weight_sgp$prior_weight) -->
<!--   prior_weight_sgp$name <- psd_vec -->
<!--   prior_weight_sgp <- rename(prior_weight_sgp, psd_sgp = name) -->
<!--   # fit the model -->
<!--   fit <- fit_sBspline_with_prior(data = data, period = fdr_df %>% filter(NAMES == gene) %>% pull(period), num_knots = k, betaprec = beta_prec, log_lib_size = log_size_vec, prior_weight = prior_weight_sgp, pred_step = 1) -->
<!--   ggplot_list[[i]] <- visualize_fit_with_prior(x = data$x, fit_result = fit) -->
<!-- } -->

<!-- # show the ggplot in a 2 x 2 grid -->
<!-- gridExtra::grid.arrange(grobs = ggplot_list, ncol = 2) -->
<!-- ``` -->


If we further zoom in the sGP group with sigma = 0:

```{r}
posterior_prob <- rowSums(posterior_matrix[, colnames(posterior_matrix) != "sgp_0"])
head(posterior_prob)
```

No genes likely fall into this group...
