---
title: "simulation_eQTL_irregular"
author: "AgueroZZ"
date: "2024-05-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
set.seed(123)
library(BayesGP)
library(TMB)
library(Matrix)
library(splines)
library(parallel)
library(ggplot2)
library(reshape2)
library(mixsqp)
library(tidyverse)
# library(mashr)
cpp_dir <- paste0(getwd(), "/code/cpp")
fig_dir <- paste0(getwd(), "/output/simulation_eQTL_irregular")
result_dir <- paste0(getwd(), "/output/simulation_eQTL_irregular")
function_dir <- paste0(getwd(), "/code/function")
source(paste0(function_dir, "/functions_fitting_Gaussian_eQTL_irregular.R"))
source(paste0(function_dir, "/functions_simulation_eQTL_irregular.R"))
compile(paste0(cpp_dir, "/Gaussian_theta_known.cpp"))
compile(paste0(cpp_dir, "/Gaussian_just_fixed.cpp"))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Gaussian_theta_known")))
dyn.load(TMB::dynlib(paste0(cpp_dir, "/Gaussian_just_fixed")))
num_cores <- detectCores() - 1
size_A = 0
size_B = 300
size_C = 4000
size_D = 700

SNR <- sample(c(10,50,100), size = (size_A + size_B + size_C + size_D), replace = TRUE)
```

```{r}
if(size_A > 0){
  data_sim_list_A <- lapply(1:size_A, function(i) simulate_process(snr = SNR[i], sd_poly = 0.2, type = "linear"))
} else{
  data_sim_list_A <- list()
}

if(size_B > 0){
  data_sim_list_B <- lapply(1:size_B, function(i) simulate_process(snr = SNR[i + size_A], sd_fun = 0.1, sd_poly = 0.2, type = "nonlinear", n_basis = 50))
} else{
  data_sim_list_B <- list()
}

if(size_C > 0){
  data_sim_list_C <- lapply(1:size_C, function(i) simulate_process(snr = SNR[i + size_A + size_B], sd_fun = 0.2, sd_poly = 0.2, type = "nonlinear", n_basis = 50))
} else{
  data_sim_list_C <- list()
}

if(size_D > 0){
  data_sim_list_D <- lapply(1:size_D, function(i) simulate_process(snr = SNR[i + size_A + size_B + size_C], sd_fun = 0.3, sd_poly = 0.2, type = "nonlinear", n_basis = 50))
} else{
  data_sim_list_D <- list()
}

```


```{r, eval=FALSE}
par(mfrow=c(2,2))
for (i in 1:4) {
  plot(data_sim_list_A[[i]]$x, data_sim_list_A[[i]]$y, 
       type = "o", col = "blue", ylim = c(-3, 3),
       lwd = 2, xlab = "Time", ylab = "estimated effect")
  lines(data_sim_list_A[[i]]$x, data_sim_list_A[[i]]$truef, col = "red", lwd = 2)
  title(paste0("Type A: Linear Dynamic eQTL ", i))
}
for (i in 1:4) {
  plot(data_sim_list_B[[i]]$x, data_sim_list_B[[i]]$y, 
       type = "o", col = "blue", ylim = c(-3, 3),
       lwd = 2, xlab = "Time", ylab = "estimated effect")
  lines(data_sim_list_B[[i]]$x, data_sim_list_B[[i]]$truef, col = "red", lwd = 2)
  title(paste0("Type B: Non-linear Dynamic eQTL ", i))
}
for (i in 1:4) {
  plot(data_sim_list_C[[i]]$x, data_sim_list_C[[i]]$y, 
       type = "o", col = "blue", ylim = c(-3, 3), 
       lwd = 2, xlab = "Time", ylab = "estimated effect")
  lines(data_sim_list_C[[i]]$x, data_sim_list_C[[i]]$truef, col = "red", lwd = 2)
  title(paste0("Type C: Non-linear Dynamic eQTL ", i))
}

for (i in 1:4) {
  plot(data_sim_list_D[[i]]$x, data_sim_list_D[[i]]$y, 
       type = "o", col = "blue", ylim = c(-3, 3), 
       lwd = 2, xlab = "Time", ylab = "estimated effect")
  lines(data_sim_list_D[[i]]$x, data_sim_list_D[[i]]$truef, col = "red", lwd = 2)
  title(paste0("Type D: Non-linear Dynamic eQTL ", i))
}
par(mfrow=c(1,1))

```

```{r}
all_datasets <- c(data_sim_list_A, data_sim_list_B, data_sim_list_C, data_sim_list_D)
```

```{r}
set.seed(123)
n_obs_vec <- sample(c(16, 8), length(all_datasets), replace = TRUE, prob = c(1/2, 1/2)) 
datasets <- list()
for (i in 1:length(all_datasets)) {
  n_obs <- n_obs_vec[i]
  # sampled_indices <- sample(1:length(all_datasets[[i]]$x), n_obs)
  if(n_obs == 16) {
    sampled_indices <- 1:16
  } else if (n_obs == 8) {
    sampled_indices <- c(1, 3, 5, 7, 9, 11, 13, 16)
  } else if (n_obs == 4) {
    sampled_indices <- c(1, 6, 11, 16)
  } else{
    stop("n_obs not supported")
  }
  datasets[[i]] <- data.frame(x = numeric(n_obs), y = numeric(n_obs), truef = numeric(n_obs), sd = all_datasets[[i]]$sd[1])
  datasets[[i]]$x <- all_datasets[[i]]$x[sampled_indices]
  datasets[[i]]$y <- all_datasets[[i]]$y[sampled_indices]
  datasets[[i]]$truef <- all_datasets[[i]]$truef[sampled_indices]
  datasets[[i]] <- arrange(datasets[[i]], by = x)
}
```

```{r}
## visualize 16 datasets
par(mfrow=c(4,4))
for (i in sample(1:length(datasets), 16)) {
  plot(all_datasets[[i]]$x, all_datasets[[i]]$truef, 
       type = "l", col = "red", ylim = range(all_datasets[[i]]$y),
       lwd = 2, xlab = "Time", ylab = "estimated effect")
  points(datasets[[i]]$x, datasets[[i]]$y, col = "blue", pch = 19)
  title(paste0("Dataset ", i))
}
par(mfrow=c(1,1))
```


To compute the SE in each dataset:
```{r}
## extract each datasets[[i]]$sd
sigma <- unlist(lapply(datasets, function(x) unique(x$sd)))
```


### **Empirical Bayes**

```{r, eval=TRUE}
set.seed(123)
p_vec <- 2
psd_iwp_vec <- sort(unique(c(0,seq(0,1, by = 0.05))))
L_vecs <- list()
# create a progress bar
pb <- txtProgressBar(min = 0, max = length(datasets), style = 3)
for (i in 1:length(datasets)) {
  setTxtProgressBar(pb, i)
  L_vecs[[i]] <- compute_log_likelihood_ospline_seq2(
    x = datasets[[i]]$x,
    y = datasets[[i]]$y,
    p = p_vec,
    num_knots = 16,
    psd_iwp_vector = psd_iwp_vec,
    pred_step = 1,
    betaprec = 0.001,
    sd_gaussian = sigma[i]
  )
}
L_matrix <- do.call(rbind, L_vecs)
save(L_matrix, file = paste0(result_dir, "/L_matrix.rda"))
```

```{r, echo=FALSE}
set.seed(123)
p_vec <- 2
psd_iwp_vec <- sort(unique(c(0,seq(0,1, by = 0.05))))
load(paste0(result_dir, "/L_matrix.rda"))
```


Then, we make use of $\texttt{mixsqp}$ to estimate the prior weights:  
```{r}
fit.sqp <- mixsqp(L = L_matrix, log = TRUE)
numiter <- nrow(fit.sqp$progress)
plot(1:numiter,fit.sqp$progress$objective,type = "b",
     pch = 20,lwd = 2,xlab = "SQP iteration",
     ylab = "objective",xaxp = c(1,numiter,numiter - 1))
prior_weight <- data.frame(p = rep(p_vec, each = length(psd_iwp_vec)), psd_iwp = psd_iwp_vec, prior_weight = fit.sqp$x)
head(prior_weight)
```


### **Posterior Inference**

With the estimated prior, we can now perform the posterior inference for each dataset:

```{r, eval=TRUE}
num_datasets <- length(datasets)
num_weights <- sum(prior_weight$prior_weight != 0)
posterior_weights_matrix <- matrix(nrow = num_datasets, ncol = num_weights)

# Loop through each dataset and perform fitting
fitted_datasets <- list()
# start a progress bar
pb <- txtProgressBar(min = 0, max = num_datasets, style = 3)
for (i in seq_along(datasets)) {
  setTxtProgressBar(pb, i)
  dataset <- datasets[[i]]
  # all_dataset <- all_datasets[[i]]
  fit_result_final <- fit_ospline_with_prior2(
    num_cores = 1,
    x = dataset$x,
    refined_x = dataset$x,
    y = dataset$y,
    num_knots = 16,
    prior_weight = prior_weight,
    betaprec = 0.001,
    sd_gaussian = sigma[i],
    pred_step = 1
  )
  fitted_datasets[[i]] <- aggregate_fit_with_prior(x = dataset$x, fit_results_with_prior = fit_result_final)$summary_df
  posterior_weights_matrix[i, ] <- fit_result_final$posterior_weights[, "posterior_weight"]
}
colnames(posterior_weights_matrix) <- paste(as.character(fit_result_final$posterior_weights[, "p"]),
                                            as.character(fit_result_final$posterior_weights[, "psd_iwp"]), sep = "_")
save(posterior_weights_matrix, file = paste0(result_dir, "/posterior_weights_matrix.rda"))
save(fitted_datasets, file = paste0(result_dir, "/fitted_datasets.rda"))
```

```{r, echo=FALSE}
num_datasets <- length(datasets)
num_weights <- sum(prior_weight$prior_weight != 0)
load(paste0(result_dir, "/posterior_weights_matrix.rda"))
load(paste0(result_dir, "/fitted_datasets.rda"))
```


We can visualize the posterior weights for each dataset:
```{r}
load(paste0(result_dir, "/posterior_weights_matrix.rda"))
posterior_weights_df <- as.data.frame(posterior_weights_matrix)
posterior_weights_df$id <- c(1:nrow(posterior_weights_matrix))
melted_data <- melt(posterior_weights_df, id.vars = "id")
melted_data$variable2 <- sub("_.*", "", melted_data$variable)
melted_data$variable3 <- as.factor(round(as.numeric(sub("*._", "", melted_data$variable)), 3))
# melted_data$variable3 <- (round(as.numeric(sub("*._", "", melted_data$variable)), 3))

ggplot(melted_data, aes(x = id, y = value, fill = variable3)) +
  geom_bar(stat = "identity") +
  labs(x = "Observation ID", y = "Weight", fill = "PSD") +
  theme_minimal() +
  # scale_fill_gradient(low = "white", high = "blue") +
  ggtitle("Structure Plot of Posterior Weights") +
  coord_flip() 
```

```{r}
## take a look at the posterior weights of observations with small number of measurements (4)
posterior_weights_df$num_measurement <- as.factor(datasets %>% map_dbl(~nrow(.)))
melted_data <- melt(posterior_weights_df, id.vars = c("id", "num_measurement"))
melted_data$variable2 <- sub("_.*", "", melted_data$variable)
melted_data$variable3 <- as.factor(round(as.numeric(sub("*._", "", melted_data$variable)), 3))
melted_data %>% filter(num_measurement == 4) %>%
  ggplot(aes(x = id, y = value, fill = variable3)) +
  geom_bar(stat = "identity") +
  labs(x = "Observation ID", y = "Weight", fill = "PSD") +
  theme_minimal() +
  ggtitle("Structure Plot of Posterior Weights for Observations with 4 Measurements") +
  coord_flip() 
```

```{r}
## take a look at the posterior weights of observations with large number of measurements (16)
melted_data %>% filter(num_measurement == 16) %>%
  ggplot(aes(x = id, y = value, fill = variable3)) +
  geom_bar(stat = "identity") +
  labs(x = "Observation ID", y = "Weight", fill = "PSD") +
  theme_minimal() +
  ggtitle("Structure Plot of Posterior Weights for Observations with 16 Measurements") +
  coord_flip() 
```


## **Effect Estimation**

Compare the posterior mean of the effect with single smoothing based on maximizing L-matrix; with MASH; with single smoothing using smoothing spline.

### **Comparison in terms of RMSE**

```{r}
fitted_datasets_ss <- list()
for (i in seq_along(datasets)) {
  dataset <- datasets[[i]]
  ss_fit <- smooth.spline(x = dataset$x, y = dataset$y)
  fitted_datasets_ss[[i]] <- data.frame(x = datasets[[i]]$x, mean = predict(ss_fit, datasets[[i]]$x)$y)
}
```

```{r, eval=TRUE}
fitted_datasets_ml <- list()
for (i in seq_along(datasets)) {
  dataset <- datasets[[i]]
  L_vec_selected <- L_matrix[i, ]
  which_sigma_max <- psd_iwp_vec[which.max(L_vec_selected)]
  fit_ospline_single <- fit_ospline(
    x = dataset$x,
    y = dataset$y,
    p = 2,
    num_knots = 16,
    psd_iwp = which_sigma_max,
    betaprec = 0.001,
    sd_gaussian = sigma[i],
    pred_step = 1, 
    refined_x = datasets[[i]]$x
  )
  fitted_datasets_ml[[i]] <- data.frame(x = all_datasets[[i]]$x, mean = apply(fit_ospline_single$samps_fitted, 1, mean))
}
save(fitted_datasets_ml, file = paste0(result_dir, "/fitted_datasets_ml.rda"))
```

```{r, echo=FALSE}
load(paste0(result_dir, "/fitted_datasets_ml.rda"))
```

```{r}
rmse_df <- data.frame(FASH = numeric(length = num_datasets), 
                      SS = numeric(length = num_datasets), 
                      ML = numeric(length = num_datasets),
                      RAW = numeric(length = num_datasets))

for (i in 1:nrow(rmse_df)) {
  rmse_df$FASH[i] <- sqrt(mean((fitted_datasets[[i]]$mean - datasets[[i]]$truef)^2))
  rmse_df$ML[i] <- sqrt(mean((fitted_datasets_ml[[i]]$mean - datasets[[i]]$truef)^2))
  rmse_df$SS[i] <- sqrt(mean((fitted_datasets_ss[[i]]$mean - datasets[[i]]$truef)^2))
  rmse_df$RAW[i] <- sqrt(mean((datasets[[i]]$y - datasets[[i]]$truef)^2))
}

# produce boxplot
rmse_df %>% ggplot() +
  geom_boxplot(aes(x = factor(1), y = FASH, fill = "FASH")) +
  geom_boxplot(aes(x = factor(2), y = SS, fill = "SS")) +
  geom_boxplot(aes(x = factor(3), y = ML, fill = "ML")) +
  geom_boxplot(aes(x = factor(4), y = RAW, fill = "RAW")) +
  labs(x = "Method", y = "RMSE") +
  theme_minimal() +
  ggtitle("Comparison of RMSE between FASH, SS and ML")


```


```{r}
# summarize into a table
rmse_df_compare <- rmse_df/sigma
rmse_df_compare$num_measurement <- as.factor(datasets %>% map_dbl(~nrow(.)))
# convert rmse to relative to sigma
rmse_df_compare %>% group_by(num_measurement) %>%
  summarise(mean(FASH), mean(SS), mean(ML), mean(RAW))
```

Plot the overall mean with respect to the num_measurement as a scatterplot

```{r}
rmse_df_compare$type <- as.factor(rep(c("A", "B", "C", "D"), c(size_A, size_B, size_C, size_D)))
rmse_df_compare$num_measurement <- (datasets %>% map_dbl(~nrow(.)))
rmse_df_compare %>% mutate(num_measurement = as.numeric(num_measurement)) %>%
  group_by(num_measurement) %>%
  summarise(mean((FASH)), mean((SS)), mean((ML))) %>%
  gather(key = "method", value = "rmse", -num_measurement) %>%
  ggplot(aes(x = num_measurement, y = rmse, color = method)) +
  geom_point() +
  geom_line() +
  labs(x = "Number of Measurements", y = "rMSE") +
  theme_minimal() +
  ggtitle("Comparison of rMSE between FASH, SS and ML")

```



```{r}
rmse_df_compare <- rmse_df/rmse_df$RAW
rmse_df_compare$type <- as.factor(rep(c("A", "B", "C", "D"), c(size_A, size_B, size_C, size_D)))
# ggplot of boxplot colored by method, facet by sigma
ggplot(melt(rmse_df_compare), aes(x = variable, y = value, fill = variable)) +
  geom_boxplot() +
  labs(x = "Method", y = "(rMSE)") +
  theme_minimal() +
  coord_cartesian(ylim = c(0, 1.5)) +
  facet_wrap(~type) +
  ggtitle("Comparison of rMSE between FASH, MASH, SS and ML")

rmse_df_compare %>% group_by(type) %>%
  summarise(mean(FASH), mean(SS), mean(ML), mean(RAW))

```




